---
title: "Project I: SEMMA with Regularized Logistic Regression"
author:
- Quaye, George Ekow
#date: "Due 9/14/2020"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
    toc: yes
    toc_depth: 4
  word_document:
    toc: yes
    toc_depth: '4'
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Data Mining}
- \lhead{SEMMA with Regularized Logistic Regression}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
geometry: margin=1in
spacing: single
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\
\newpage
\section {Bringing in the data}
**1. Bring the data into R (or Python)**
\
```{r}
#Reading the data
data <- read.table(file = "diabetes_data_upload.csv",sep=",", header = T, na.strings = c("NA", "", " "), 
                   stringsAsFactors = T)
dim(data)
```
\
The data set has 520 observations and 17 columns.       

\section{Exploratory Data Analysis}
\subsection{Variable Type}           
```{r}
colnames(data)
```
\
```{r}
str(data)
```
\
A data with 520 observations on 17 variables, 1 being numeric variables and 16 being nominal(categorical) variables.
\
\subsection{Frequency distribution of the target variable}                   

**inspect the frequency distribution of the target variable class and see, e.g., whether we
have an unbalanced classification problem.**      
\
\
```{r}
#inspect the frequency distribution of the target variable class.
library(questionr)
freq(data$class, total=T)
```
\
From the output the counts for negative are 200 and that for positive are 320. Thus in this scenario, we don’t have a very unbalanced classification problem.
\
\subsection{Checking for Missing values}
**Are there missing values? If so, handle them with an appropriate strategy such as listwise deletion or single/multiple imputation.**
```{r}
# INSPECT THE DISTINCT VALUES OF EACH X
cols<- 2:NCOL(data)
for (j in cols){
 print(colnames(data)[j])
 print(table(table(data[,j]), useNA="ifany"))
}
```
\
```{r}
# MISSING PERCENTAGES FOR ALL COLUMNS (OR VARIABLES)
colMeans(is.na(data))
```
\
There are no missing values in the data set.      

\section{Variable Screening}
\subsection{Chi-Square test of association}
**Explore the marginal (bivariate) associations between class and each attribute/predictor.**
```{r , message=FALSE}
#Bivariate association of the response with the categorical predictors.
m<-data[,-c(1)]

library(car);
cols.x <- 1:(NCOL(m)-1)
xnames <- names(m)[cols.x]
y <- m$class
OUT <- NULL
for (j in 1:length(cols.x)){
x <- data[, cols.x[j]]
xname <- xnames[j]
tbl <- table(y, x)
pvalue <- chisq.test(tbl)$p.value
OUT <- rbind(OUT, cbind(xname=xname, pvalue=pvalue)) }
```
\
```{r}
OUT <- as.data.frame(OUT) 
colnames(OUT) <- c("name", "pvalue") 
OUT
```
\
\subsection{Deletion of non significant variables}
```{r}
#Taking the non significant variables out
data<-data[,-c(11,13)]
data$class<- ifelse(data$class=="Negative", 0,1)
colnames(data)
```
\
From the output, all the predictors are significant except partial.paresis and irritability given the threshold probability of $0.25$. Thus there is significant evidence there exist an association between Class and all significant attributes.
\
\subsection{Correlation plot among the variables}
```{r}
library(GoodmanKruskal)
data1<- data[,-c(17)]
dat<- GKtauDataframe(data1)
plot(dat, corColors = "magenta")
```
\
There appear to be  no correlation between the predictor variables             

\section{Data Partition}
**4) Partition the data into two parts, the training data D1 and the test data D2, with a ratio of 2:1.**
\
```{r}
#Partitioning data
set.seed(125)
sampleData <- sample(nrow(data), (2.0/3.0)*nrow(data), replace = FALSE) # training set
D1 <- data[sampleData, ]
# test set
D2 <- data[-sampleData, ]
dim(D1)
```
\
Given the train data there are 346 observations with 15 variables.
\
```{r}
dim(D2)
```
\
Given the test data there are 174 observations with 15 variables.           

\section{Logistic Regression Modeling}
\subsection{Fitting the model}
**5)a We now build a logistic regression model for this medical diagnosis task.**
\
```{r , message=FALSE}
#Fitting a regularized logistic regression model
library(ncvreg)
library(glmnet)
formula0 <- class~Age + Gender + Polyuria + Polydipsia + sudden.weight.loss + weakness + Polyphagia + Genital.thrush + visual.blurring + Itching + delayed.healing + muscle.stiffness + Alopecia + Obesity

X <- model.matrix(as.formula(formula0), data=D1)
y <- D1$class
XTest <- model.matrix(as.formula(formula0), data=D2)
ytest <- D2$class

library(verification)
Lambda <- seq(0.0001, 0.5, length.out = 500)
L <- length(Lambda)
OUT <- matrix(0, L, 4)
for (i in 1:L){
    fit.lasso <- glmnet(x=X, y=y, family="binomial", alpha=1, # LASSO
	    lambda = Lambda[i], standardize=T, thresh = 1e-07, 
      	maxit=3000)
    pred <- predict(fit.lasso, newx=XTest, s=Lambda[i], type="response")
    missRate <- mean(ytest != (pred > 0.5))
    mse <- mean((ytest-pred)^2)
    AUC <- roc.area(obs=ytest, pred=pred)$A
    OUT[i, ] <- c(Lambda[i], missRate, mse, AUC)
}
head(OUT)

par(mfrow = c(1,2))
plot(OUT[,1], OUT[,2], type = "b", col="blue",ylab = "Missclassification Rate", xlab = expression(lambda))
plot(OUT[,1], OUT[,3], type = "b", col="red",ylab = "MSE", xlab = expression(lambda))

```
\
Given the plot of missclassifcation rate it is seen that as $\lambda$ increases, the classification rate also increases. However when $\lambda >= 0.2$, the missclassification rate remains constant. Also given the plot of MSE it is noticed that as $\lambda$ increases, the MSE also increases. However when $\lambda >= 0.3$, the MSE remains constant.                 
\
\subsection{Selecting best tuning parameter using validation data}
\
```{r}
#Selection of tuning parameter using the validation data D2
lambda.best <- OUT[which.min(OUT[,3]), 1]; lambda.best
```
\
The criteria used to select the tuning parameter is the mean square error for the predicted probabilities.
\          
\subsection{Final best model fit}
**b) Present your final `best' model fit. Which variables are important predictors? Interpret the results.**
\
```{r}
Xnew <- rbind(X, XTest)
ynew <- c(y, ytest)
fit.best <- glmnet(x=Xnew, y=ynew, family="binomial", alpha=1, 
	    lambda = lambda.best, standardize=T, thresh = 1e-07,maxit=3000)
```
\
        
\subsection{Checking for important predictors}
```{r}
#Checking for important predictors.
fit.best$beta
```
\
From the output, the coefficients with non zero values are the important predictors. 
\

\section{Model Assestment / Deployment}
6) Apply the final logistic model to the test data $D2$. Present the ROC curve and the area under the curve, i.e., the C-index.     

```{r}
# Final model to test data
FinalPred <- predict(fit.best, newx=XTest, s=lambda.best, type="response")
```
\
```{r, message=FALSE}
# ROC Curve and AUC
library(cvAUC)
AUC <- ci.cvAUC(predictions=FinalPred, labels=ytest, folds=1:NROW(D2), confidence=0.95); AUC 
auc.ci <- round(AUC$ci, digits=3)

mod.glm <- verify(obs=ytest, pred=FinalPred)
roc.plot(mod.glm, plot.thres = NULL)
text(x=0.7, y=0.2, paste("Area under ROC =", round(AUC$cvAUC, digits=3),
	"with 95% CI (", auc.ci[1], ",", auc.ci[2], ").",
	sep=" "), col="blue", cex=1.2)
```
\
The Area under ROC curve is obtained as $0.986$ and the confidence interval for the area under ROC curve is also shown on the plot with 95% confidence level.


